\papertitle

### 2025-04-30 paper/main.tex
update paper

% import encoding and font packages for pdflatex, in order
\usepackage[T1]{fontenc} % output T1 font encoding (8-bit) so accented characters are a single glyph
\usepackage[utf8]{inputenc} % allow input of utf-8 encoded characters
\usepackage{ebgaramond} % document's serif font
\usepackage{tgheros} % document's sans serif font

% import language, regionalization, microtype, and setspace in order
\usepackage[strict,autostyle]{csquotes} % smart and nestable quote marks
\usepackage[USenglish]{babel} % automatically regionalize hyphens, quote marks, etc
\usepackage{microtype} % improves text appearance with kerning, etc
\usepackage{setspace} % configure spacing between lines
\usepackage{mathtools} % use package math tool
% import everything else
\usepackage{abstract} % allow full-page title/abstract in twocolumn mode
\usepackage{amsmath}
\usepackage{authblk} % footnote-style author/affiliation info
\usepackage{booktabs} % better looking tables
\usepackage{caption} % custom figure/table caption styles
\usepackage{datetime} % enable formatting of date output
\usepackage[final]{draftwatermark} % watermark paper as a draft
\usepackage{endnotes} % enable endnotes
\usepackage{geometry} % configure page dimensions and margins
\usepackage{graphicx} % better inclusion of graphics
\usepackage{hyperref} % hypertext in document
\usepackage{natbib} % author-year citations w/ bibtex, including textual and parenthetical
\usepackage{rotating} % rotate wide tables or figures on a page to make them landscape
\usepackage{titlesec} % custom section and subsection heading
\usepackage{url} % make nice line-breakble urls
\usepackage{makecell}

@@

% location of figure files, via graphicx package
\graphicspath{{./figures/}}

@@

\title{\papertitle}
\author[]{Redacted for review}%{\myname}
\affil[]{Redacted for review}%{\myaffiliation}
\date{}%{\paperdate}

@@

Travel time prediction is central to accessibility analyses, sustainable transportation infrastructure provision, and active transportation interventions. However, calculating accurate travel times, especially for driving, requires either extensive technical capacities and bespoke data, or sources like the Google Maps API that quickly becomes prohibitively expensive to analyze thousands or millions of trips necessary for metropolitan-scale planning. These obstacles make it difficult for researchers, practitioners, and community advocates to develop an evidence base for more sustainable transportation and equitable accessibility. This article develops a free, open-source minimally-congested driving time prediction model with minimal cost, data, and computational requirements. It trains and tests this model using the Los Angeles, California urban area as a case study by calculating naïve travel times then developing a predictive random forest regression model of Google Maps' \enquote{gold standard} travel times using those naïve times plus turns and traffic controls along the naïve routes. Validation reveals that this method yields accurate, free, and simple travel time predictions.

@@

Travel time prediction underpins many transport planning processes and research problems. It is key to understanding urban accessibility, transport mode and route choice, and location decisions. Predicting congested travel times (i.e., when traffic congestion impedes flows) requires large volumes of real-time data on traffic conditions. Such data are proprietary, expensive, and beyond the reach of many planning practitioners and scholars. This is usually not a problem for well-resourced organizations with highly-skilled labor. However, technical expertise, complex data and software requirements, and high costs pose insurmountable obstacles for many urban planning researchers, practitioners, and community advocates.

Although typical \enquote{naïve} travel time prediction methods (which minimize travel distance or street segment traversal time) provide easy solutions, they ignore delays from turns and traffic controls and thus usually under-predict real-world driving travel time. However, minimally-congested travel time prediction offers a middle ground of more-accurate predictions than naïve travel times do---yet much lower data and computing requirements than traffic-aware and congested travel time prediction.

To address this need for a middle ground, this article contributes a simple, accurate, and free method of minimally-congested driving travel time prediction. Our results show a substantial improvement over traditional, naïve methods of travel time prediction common in the urban planning literature. As we shall show in this article, across the Los Angeles urban area, our average trip's travel time prediction differs on average from Google's \enquote{gold standard} prediction by only 0.38 seconds. In contrast, simple---but common---naïve travel time prediction minimizing max-speed edge traversal times under-predicts Google's travel times by 182.71 seconds, on average.

@@

Recent studies in the computer science and engineering literatures propose a variety of prediction techniques with heavy computing and data requirements. They usually measure their prediction accuracy by mean absolute percentage error (MAPE), defined in Equation~\ref{eq:mape}:

@@

Finally, many planning scholars rely instead on the Google Maps API to obtain accurate origin-destination travel times and shortest paths \citep[e.g.,][]{fielbaum2021assessment,costa2021spatial, swayne2021integrating,hu2020estimating,cuervo2022dynamic,chen2020communities,hwang2024measuring}. As mentioned previously, Google travel times represent something of a \enquote{gold standard} in the literature, but come with drawbacks: the algorithms are closed-source and---for large batches of queries, such as for simulating metropolitan-scale trip taking---it can become prohibitively expensive and require many API queries. Nevertheless, evidence shows Google travel times offer a good approximation of real-world travel times \citep{lin2021impact,fu2023comparative,alsobky2020estimating,wang2011estimating}.

@@

A better way would be to seek a middle ground between expensive, complicated, point-to-point empirical travel times and overly-simple naïve predictions. Such free empirical travel time predictions may not reflect congested travel times (which vary drastically throughout the day and week), but would provide a good prediction of minimally-congested travel times. This middle-ground method should meet three criteria. First, it should only use free software and free data, avoiding expensive and massive Google travel time queries or bespoke GPS data collection. Second, it should be a simple and accessible method for urban planners without advanced technical skills. Third, it should provide a substantial improvement in accuracy or spatial resolution over naïve predictions such as minimizing Euclidean distance, network distance, street edge traversal time, and exsiting TAZ-to-TAZ travel times, and approach the accuracy of the state-of-art models. Meeting these three criteria, a middle ground should be inexpensive, easy-to-use, and reasonably accurate.

@@

This article proposes a new generalizable method to solve this problem. In short, it collects a small free amount of data from the Google Maps API plus free open data to train a local model to predict accurate minimally-congested travel times using relatively simple standard tools and techniques. Figure~\ref{fig:workflow} shows a detailed workflow.

\begin{figure*}[hbt!]
    \centering
    \includegraphics[width=1.0\textwidth]{fig_workflow.jpg}
    \caption{Detailed model workflow.}\label{fig:workflow}
\end{figure*}

\begin{table*}[hbt!]
    \centering
    \caption{Traffic control elements in the network model, as identified by OSM tags.}\label{tab:node_count}
    \begin{tabular}{lr}
        \toprule
        Traffic control element    & Node count \\ \midrule
        Crossing                   &     21,559 \\
        Stop sign                  &     26,304 \\
        Turning circle             &     17,886 \\
        Traffic signal             &     15,261 \\
        Motorway junction          &      1,436 \\
        Mini roundabout            &         44 \\
        Give way                   &        189 \\
        3-way+ street intersection &    126,982 \\
        Nodes without any tag      &    699,912 \\
        Total nodes                &    782,825 \\ \bottomrule
    \end{tabular}
\end{table*}

@@

We define the study area as the convex hull around the intersection of the Los Angeles County boundary and the Los Angeles urban area boundary from the Global Human Settlement Layer's Urban Center Database \citep{florczyk2019description, GHS2019}. This allows us to retain the main urbanized area without adjacent metropolitan areas (such as the Inland Empire). We then model the drivable street network within this study area from OpenStreetMap using the OSMnx software \citep{boeing_modeling_2025}, retaining the strongly connected component, to generate a graph with 782,825 nodes. This graph is unsimplified, meaning that nodes represent intersections and dead-ends, as well as curving street line geometry vertices. Table~\ref{tab:node_count} summarizes these nodes' tagged traffic control elements: these data are sparse and most intersections lack traffic control information.

To generate origin-destination (OD) pairs for training a prediction model, we first over-sample 5,000,000 random node pairs from the street intersections and dead-ends in this graph. We then filter these down to realistic, minimally-congested trip patterns using the most-recently released Uber movement data \citep{ubermovement2020}. These data derive from from Uber trip GPS traces aggregated by tract-to-tract flow and hour of the day. We filter our OD pairs down to those that have a matching real-world trip ($n$ = 1,197,513) that occurred during the 03:00 hour ($n$ = 41,360) to best approximate minimally-congested traffic conditions.

Finally, we collect \enquote{gold standard} travel times from the Google Maps API (specifically, the Routes API), which has been used in many studies to ground-truth travel times  \citep[e.g.,][]{ludwig2023traffic, hu2020estimating, wang2011estimating, fu2023comparative, delmelle2019travel}. For each OD pair, the API provides the fastest network path, its travel time, and its length. The API allows users to predict travel times for trips departing in the future: the closer the departure time, the more accurate the prediction. Accordingly, we set the departure time to 03:00 and performed the query immediately beforehand (between 02:30--02:40) on 31 January and 1 February 2024. We used Google's \enquote{BEST\_GUESS} traffic-aware model to predict its \enquote{duration\_in\_traffic} travel time. The API was unable to solve 16 OD pairs' routes, resulting in 41,360 OD pairs with a Google travel time, which we used for prediction modeling.

\begin{figure*}[hbt!]
    \centering
    \includegraphics[width=1.0\textwidth]{fig_turns_definition.jpg}
    \caption{Angular definition of turns in the model.}\label{fig:turns_definition}
\end{figure*}

\subsection{Travel Time Prediction}

Our travel time prediction model has two steps. First, we calculate \enquote{naïve travel times} across a set of trips using open data and open-source software. Second, we train a random forest regression model to predict the Google \enquote{gold standard} travel times from our \enquote{naïve travel times} plus a set of covariates. This subsection details this process.

To calculate naïve travel times, we solve each OD pair's shortest path using Dijkstra's algorithm to minimize edge traversal time at the speed limit. Then we count the number of traffic control elements (including traffic signals, pedestrian crossings, give ways, and mini roundabouts) and turns (i.e., left, slight left, right, slight right, and u-turn), as shown in Figure~\ref{fig:turns_definition}, along each OD pair's route.

Then we train a random forest regression model to predict a route's Google travel time from  the naïve travel time plus the count of each type of traffic control element and turn encountered along the naïve shortest path. We also tested other algorithms, including ordinary least squares, gradient boosting regression, AdaBoost regression, and decision tree regression, but settled on random forest regression as its ensemble accuracy substantially outperforms the others.

We divide our OD pairs into a standard 80/20 training/test split, then tune the hyperparameters through a randomized grid search with 5-fold cross-validation by calculating their mean absolute error (MAE), defined in Equation~\ref{eq:mae}:

\begin{equation}
\label{eq:mae}
\text{MAE} = \frac{\sum^{n}_{i=1} \left|{t_i - \hat{t_i}}\right|}{n}
\end{equation}

where $n$ is the total number of trips, $t_i$ is trip $i$'s observed travel time, and $\hat{t_i}$ is trip $i$'s predicted travel time.

\subsection{Validation}

We validate our out-of-sample predictions against the \enquote{gold standard} Google travel times using five measures. First, we calculate the MAPE, following Equation~\ref{eq:mape}. Second, we calculate the MAE, following Equation~\ref{eq:mae}, to measure the average absolute difference in seconds between our prediction and the reference Google value. Third, we perform difference-in-means ($\delta$) $t$-tests to determine whether our predicted travel times are statistically significantly different (at the $p<=0.05$ significance level) from the reference Google values. Despite the large sample size, we expect insignificant $t$-statistics if the predicted travel times' distribution approximates the Google travel times' distribution. Fourth, we calculate the standard $R^2$ to measure our model's success in explaining travel time variation. Fifth and finally, we calculate the average of the pairwise ratios (APR) of our predicted travel time and the Google travel time to measure the average over- or under-prediction of the trips' travel times, as defined in Equation~\ref{eq:apr}:

\begin{equation}
\label{eq:apr}
\text{APR} = \frac{\sum^{n}_{i=1}\frac{t_i}{\hat{t_i}}}{n}
\end{equation}

where $n$ is the total trip count, $t_i$ is trip $i$'s Google travel time, and $\hat{t_i}$ is our predicted travel time for trip $i$.

@@

\subsection{Optimized Hyperparameters}

Our final optimized model hyperparameterization uses 600 decision trees, random sampling with replacement in each decision tree, a maximum decision tree depth of 10, a non-limited maximum number of input explanatory features, a requirement of at least two samples to split at a decision node and one sample at a leaf node, and equal sample weighting. Finally, we check our trained model for overfitting by conducting 5-fold cross-validation across the whole sample using tuned hyperparameters: the resulting five MAE values (75.3, 74.5, 73.5, 73.3, 74.7) are all similar to each other, which indicates no overfitting problem as the hyperparameters are not overly specific to the training set.

@@

Table~\ref{tab:validation_results} details our travel time prediction model's out-of-sample improvement over the initial naïve travel time calculation. Our model achieves a MAPE of 8.4\%, in line with the approximately 10\% MAPEs seen in the state-of-the-art travel time prediction literature, but without their extensive and expensive input data requirements. In comparison, our initial naïve travel time calculation yields a much worse MAPE of 21.1\%. Whereas the naïve travel time calculation's MAE is 183.5 seconds from the Google travel time, our predicted travel times' MAE is just 75.2 seconds---an improvement by a factor of 2.4.

\begin{table*}[tb!]
    \centering
    \caption{Results of out-of-sample (testing set) prediction of (1) our prediction model and (2) the initial naïve travel time calculation, validated against the corresponding Google travel times. The five validation measures are (1) MAPE, (2) MAE, (3) difference-in-means ($\delta$) and corresponding $t$-test $p$-value, (4) APR, and (5) $R^2$. See methods for definitions.}\label{tab:validation_results}
    \begin{tabular}{lrr}
        \toprule
                                      & Prediction model & Naïve calculation \\
        \midrule
        MAPE (\%)                     &             8.43 &             21.13 \\
        MAE (seconds)                 &            75.23 &            183.55 \\
        $\delta$ (seconds)            &             0.38 &           -182.71 \\
        $p$-value                     &             0.75 &             <0.01 \\
        APR                           &             1.01 &              0.79 \\
        $R^2$                         &             0.93 &              0.74 \\
        \midrule
        $n$                           &           8,272 &            8,272 \\
        \bottomrule
    \end{tabular}
\end{table*}

The $t$-test reveals significant differences between the initial naïve travel time calculations and the Google travel times: the $\delta$ of -182.7 seconds corresponds to $p<0.01$. In other words, the naïve travel time under-predicts Google by over 3 minutes. However, the $t$-test reveals insignificant differences between our model's out-of-sample predictions and the Google travel times: the $\delta$ of 0.4 seconds corresponds to a $p$ value of 0.75. That is, our prediction model over-predicts by less than half a second on average, which is statistically insignificantly different from zero. If we compare the aforementioned MAEs to these differences-in-means, we can see that the initial naïve travel time calculations significantly and consistently under-predict travel time, but our prediction model shows no such directional bias: its random error balances out between (smaller absolute) over-and under-prediction.

The APR offers another lens on this finding. On average across the OD pairs, the initial naïve travel time calculation under-predicts Google travel time by 21\%, but our prediction model over-predicts it by just 1\%. Our model also explains more of the variation in travel time: its $R^2$ of 0.93 is substantially higher than the 0.74 $R^2$ of the initial naïve travel time calculation. In summary, each of our five validation measures demonstrates that our prediction model drastically improves on naïve travel time calculations---without needing extensive, expensive input data or advanced deep learning software---and offers travel time predictions much closer to those of Google's \enquote{gold standard} predictions.

@@

Re-state the problem. This study set out to investigate if a middle-ground travel time prediction model was possible. It identified three criteria for success: (1) it should use free software and data and avoid bespoke sensor data collection; (2) it should be easy and accessible to use; (3) it should offer better accuracy than naïve predictions (e.g., minimizing Euclidean distance, network distance, street edge traversal time) and better spatial resolution than zone-to-zone travel times.

@@

This middle ground is not necessarily for computer scientists, high-end labs with top-of-the-line computing hardware, or well-resourced agencies with extensive data budgets. Rather, it empowers the scholars and practitioners who need it the most---those working on the frontlines of urban planning and policymaking. These scholars and practitioners rely on simple but inaccurate routing models---such as minimizing Euclidean distance, network distance, or edge traversal time---when they lack the technical expertise, computing equipment, or extensive data needed to implement the literature's cutting-edge models. However, this study demonstrated how these naïve methods systematically under-predict real-world travel times. Our model addresses this problem by incorporating sparse open data on traffic control elements and turns into the prediction and achieves an accuracy in line with state-of-the-art methods---but without their data needs or computing requirements. In other words, our model does not replace the state-of-the-art methods for those with extensive resources available to implement them, but rather offers far more accurate predictions for urban planners who would otherwise fall back on simpler naïve models. Inaccurate travel times skew empirical understandings and mislead planning interventions. Our model represents an important contribution for urban planning scholarship and evidence-based practitioner interventions.

>>>>

% load encoding and font packages for pdflatex, in order
\usepackage[T1]{fontenc}    % output 8-bit encoded fonts
\usepackage[utf8]{inputenc} % allow input of utf-8 encoded characters
\usepackage{ebgaramond}     % document's serif font
\usepackage{tgheros}        % document's sans serif font

% load babel, csquotes, and microtype in order
\usepackage[USenglish]{babel} % auto-regionalize hyphens, quote marks, etc
\usepackage[strict,autostyle]{csquotes} % smart and nestable quote marks
\usepackage[babel=true]{microtype} % enable micro-typographic adjustments

% import everything else
\usepackage{amsmath} % additional mathematical typesetting features
\usepackage{authblk} % footnote-style author/affiliation info
\usepackage{booktabs} % better looking tables
\usepackage{caption} % custom figure/table caption styles
\usepackage{datetime} % enable formatting of date output
\usepackage[final]{draftwatermark} % watermark paper as a draft
\usepackage{endnotes} % enable endnotes
\usepackage{geometry} % configure page dimensions and margins
\usepackage{graphicx} % better inclusion of graphics
\usepackage{natbib} % author-year citations w/ bibtex, including textual and parenthetical
\usepackage{rotating} % rotate wide tables or figures on a page to make them landscape
\usepackage{setspace} % configure spacing between lines
\usepackage{titlesec} % custom section and subsection heading
\usepackage{url} % make nice line-breakble urls

% load hyperref/orcidlink last for compatibility
\usepackage{hyperref} % enable hyperlinks and pdf metadata
\usepackage{orcidlink} % provide orcid logo and link

@@

% location of figure files, via graphicx package
\graphicspath{{.}}

@@

\title{\papertitle}
\author[]{Redacted for review}%{\myname}
\affil[]{Redacted for review}%{\myaffiliation}
\date{}%\paperdate}

@@

Travel time prediction is central to accessibility analyses, sustainable transportation infrastructure provision, and active transportation interventions. However, calculating accurate travel times, especially for driving, requires either extensive technical capacity and bespoke data, or sources like the Google Maps API that quickly become prohibitively expensive to analyze thousands or millions of trips necessary for metropolitan-scale planning. These obstacles make it difficult for researchers, practitioners, and community advocates to develop an evidence base for more sustainable transportation and equitable accessibility. This article introduces a free, open-source minimally-congested driving time prediction model with minimal cost, data, and computational requirements. It trains and tests this model using the Los Angeles, California urban area as a case study by calculating naïve travel times from open data then developing a random forest model to predict true travel times as a function of those naïve times plus open data on turns and traffic controls. Validation shows that this method produces accurate, free, and simple travel time predictions, offering a superior middle-ground technique for urban planners.

@@

Travel time prediction underpins many transport planning processes and research problems. It is key to understanding urban accessibility, transport mode and route choice, and location decisions \citep{zhang2015gradient,jenelius2013travel}. Predicting congested travel times (i.e., when traffic congestion impedes flows) requires large volumes of real-time data on traffic conditions \citep[e.g.,][]{hou2018network}. Such data are proprietary, expensive, and beyond the reach of many planning practitioners and scholars \citep{carrion2012value,wang2011estimating}. This is usually not a problem for well-resourced organizations with highly-skilled labor---however, technical expertise, complex data and software requirements, and high costs pose insurmountable obstacles for many urban planning researchers, practitioners, and community advocates \citep{giles2022creating}.

Although typical \enquote{naïve} travel time prediction methods (which minimize travel distance or street segment traversal time) provide easy solutions, they ignore delays from turns and traffic controls and thus usually under-predict real-world driving travel time \citep{ludwig2023traffic,salonen2013modelling,yiannakoulias2013estimating}. However, as we will argue, minimally-congested travel time prediction offers a better middle ground of more-accurate predictions than naïve travel times do---yet much lower data and computing requirements than real-time traffic-aware congested travel time prediction.

To address this need for a middle ground, this article contributes a novel accurate, easy-to-use, and free method of minimally-congested driving travel time prediction. Our results show a substantial improvement over traditional, naïve methods of travel time prediction common in the urban planning literature. As we shall show in this article, across the Los Angeles urban area, our average trip's travel time prediction differs from Google's \enquote{gold standard} prediction by only 0.45 seconds. In contrast, simple---but common---naïve travel time prediction minimizing max-speed edge traversal times under-predicts by 183.55 seconds on average.

@@

Recent studies in the computer science and engineering literatures propose a variety of prediction techniques with heavy computing and data requirements. They usually measure their prediction accuracy by mean absolute percentage error (MAPE), defined in \autoref{eq:mape}:

@@

Finally, many planning scholars rely instead on the Google Maps API to obtain accurate origin-destination travel times and shortest paths \citep[e.g.,][]{fielbaum2021assessment,costa2021spatial, swayne2021integrating,hu2020estimating,cuervo2022dynamic,chen2020communities,hwang2024measuring}. As mentioned previously, Google travel times represent something of a \enquote{gold standard} in the literature, but come with drawbacks: the algorithms are closed-source and---for large batches of queries, such as for simulating metropolitan-scale trip taking---it can become prohibitively expensive and require many API queries. Nevertheless, evidence shows Google travel times offer a good proxy for observed real-world travel times \citep{lin2021impact,fu2023comparative,alsobky2020estimating,wang2011estimating}.

@@

A better way would be to seek a middle ground between expensive, complicated, point-to-point empirical travel times and overly-simple naïve predictions. Such free empirical travel time predictions may not reflect congested travel times (which vary drastically throughout the day and week), but would provide a good prediction of minimally-congested travel times. This middle-ground method should meet three criteria. First, it should only use free software and free data, avoiding expensive API queries or bespoke GPS data collection. Second, it should be a simple and accessible method for urban planners without advanced technical skills. Third, it should provide a substantial improvement in accuracy or spatial resolution over naïve predictions such as minimizing Euclidean distance, network distance, street edge traversal time, and exsiting TAZ-to-TAZ travel times, and approach the accuracy of the state-of-art models. Meeting these three criteria, a middle ground should be inexpensive, easy-to-use, and reasonably accurate.

@@

This article proposes a novel open-source, reusable, generalizable method\endnote{Our open-source tool is hosted at [REDACTED] for others to freely reuse.} to solve this problem. In short, it collects free open data plus a small amount of free data from the Google Maps API to train a local model to predict minimally-congested driving travel times with sparse data and simple computing hardware. \autoref{fig:workflow} shows a detailed workflow.

\begin{figure*}[bt!]
    \centering
    \includegraphics[width=1.0\textwidth]{fig_workflow.jpg}
    \caption{Detailed model workflow.}\label{fig:workflow}
\end{figure*}

\begin{table*}[tb!]
    \centering
    \caption{Traffic control elements (as identified by OpenStreetMap tags) present in the graph, demonstrating their sparseness relative to the total intersections and nodes in the graph.}\label{tab:traffic_control_counts}
    \begin{tabular}{lr}
        \toprule
        Element                        &   Count \\
        \midrule
        Crossing                       &  21,560 \\
        Stop sign                      &  26,304 \\
        Traffic signal                 &  15,262 \\
        Mini roundabout                &      44 \\
        Give way                       &     189 \\
        \midrule
        Total traffic control elements &  63,359 \\
        Total street intersections     & 127,093 \\
        Total nodes                    & 782,825 \\
        \bottomrule
    \end{tabular}
\end{table*}

@@

We define an implementation study area as the convex hull around the intersection of the Los Angeles County boundary and the Los Angeles urban area boundary from the Global Human Settlement Layer's Urban Center Database \citep{florczyk2019description, GHS2019}. This allows us to retain the main urbanized area without adjacent metropolitan areas (such as the Inland Empire). We then model the drivable street network within this study area from OpenStreetMap using the OSMnx package \citep{boeing_modeling_2025} and retaining the strongly connected component, to produce an unsimplified graph with 782,825 nodes, of which 127,093 are street intersections. The graph also contains 63,359 tagged traffic control elements. \autoref{tab:traffic_control_counts} summarizes these nodes' tagged traffic controls: these data are sparse and many true intersections lack traffic control information.

To generate origin-destination (OD) pairs for training a prediction model, we first over-sample 5,000,000 random node pairs from the street intersections and dead-ends in this graph. We then filter these down to realistic, minimally-congested trip patterns by using the most-recently released Uber movement data \citep{ubermovement2020}. These data derive from Uber trip GPS traces aggregated by tract-to-tract flow and hour of the day. We filter our OD pairs down to those that have a matching real-world trip ($n$ = 1,197,513) that occurred during the 03:00 hour ($n$ = 41,378) to best approximate minimally-congested traffic conditions. These Uber data are not a perfect measure of all real-world trips, but do validate the presence of travel demand between different OD pairs at different times of day.

Then we proxy \enquote{true travel times} by collecting travel times from the Google Maps Routes API, which has been used in numerous studies as a \enquote{gold standard} to proxy real-world travel times as discussed in the background section \citep[e.g.,][]{ludwig2023traffic, hu2020estimating, wang2011estimating, fu2023comparative, delmelle2019travel}. For each OD pair, the API provides the fastest network path, its travel time, and its length. The API allows users to predict travel times for trips departing in the future: the closer the departure time, the more accurate the prediction. Accordingly, we set the departure time to 03:00 and performed the query immediately beforehand (between 02:30--02:40) on 31 January and 1 February 2024. We used Google's \enquote{BEST\_GUESS} traffic-aware model to predict its \enquote{duration\_in\_traffic} travel time. The API was unable to solve 18 OD pairs' routes, resulting in 41,360 OD pairs with a true travel time, which we use as the response in our prediction model.

\begin{figure*}[tb!]
    \centering
    \includegraphics[width=1.0\textwidth]{fig_turns_definition.jpg}
    \caption{Angular definition of turns in the prediction model.}\label{fig:turns_definition}
\end{figure*}

\subsection{Model Specification and Tuning}

Conceptually, our travel time prediction model has two steps. First, we calculate \enquote{naïve} travel times across a set of trips using open data and open-source software. Second, we train a random forest (regression) model to predict our true travel times from our naïve travel times plus a set of covariates.

To calculate the naïve travel times for step one, we solve each OD pair's shortest path using Dijkstra's algorithm to minimize edge traversal time at the speed limit. Then we count the number of traffic control elements across the 5 types in \autoref{tab:traffic_control_counts} (stop signs, traffic signals, pedestrian crossings, give ways, and mini roundabouts), and the number of turns across the 5 types in \autoref{fig:turns_definition} (left, slight left, right, slight right, and u-turn), that were encountered along each OD pair's naïve route. Next we specify a travel time prediction model to predict an OD pair's true travel time (proxied by Google travel time) as a function of its naïve travel time plus these counts of each type of traffic control element. This model's specification is generalized by \autoref{eq:prediction_model}:

\begin{equation}
\label{eq:prediction_model}
y = f(X) + \epsilon
\end{equation}

where $f$ is a prediction model to be trained on our data, $y$ is a length-$n$ response vector (true travel time), $X$ is a matrix of $n$ observations on 11 predictors (naïve travel time plus counts of the 5 types of traffic control element encountered and counts of the 5 types of turn directions encountered), and $\epsilon$ is random error.

We train candidate models of this specification using several algorithms: decision trees, random forest, gradient boosting, and AdaBoost. A set of hyperparameters controls this algorithmic training. To tune each's hyperparameters for optimal performance, we split our data into a standard 80/20 training/test split then conduct a randomized grid search with 5-fold cross-validation to minimize the mean absolute error (MAE), defined by \autoref{eq:mae}:

\begin{equation}
\label{eq:mae}
\text{MAE} = \frac{\sum^{n}_{i=1} \left|{y_i - \hat{y_i}}\right|}{n}
\end{equation}

where $n$ is the total trip count, $y_i$ is trip $i$'s true travel time, and $\hat{y_i}$ is $i$'s predicted travel time.

\subsection{Model Selection and Validation}

Once we have our candidate models tuned and trained, we select our final prediction model and validate the out-of-sample predictions against true travel times via six accuracy indicators. First, we calculate the MAPE, following \autoref{eq:mape}. Second, we calculate the MAE, following \autoref{eq:mae}. Third, we calculate the mean squared error (MSE), which is more sensitive to outliers than MAE or MAPE, following \autoref{eq:mse} (with the same variables as defined for \autoref{eq:mae}):

\begin{equation}
\label{eq:mse}
\text{MSE} = \frac{\sum^{n}_{i=1} (y_i - \hat{y_i})^2}{n}
\end{equation}

Fourth, we perform difference-in-means ($\delta$) $t$-tests to determine whether our predicted travel times are statistically significantly different (at the 95\% confidence level) from the true travel times, to detect absolute systematic prediction bias. Despite the large sample size, we expect insignificant $t$-statistics if the predicted travel times' distribution closely matches the true travel times' distribution. Fifth, we calculate the average pairwise ratio (APR) of our predicted travel times and true travel times, to detect relative systematic prediction bias, as defined by \autoref{eq:apr} (with the same variables as defined for \autoref{eq:mae}):

\begin{equation}
\label{eq:apr}
\text{APR} = \frac{\sum^{n}_{i=1}\frac{y_i}{\hat{y_i}}}{n}
\end{equation}

Sixth and finally, we calculate the coefficient of determination, $R^2$, to measure our model's ability to explain observed travel time variation.

@@

\subsection{Optimized Hyperparameterization}

After tuning, the four model training algorithms perform similarly across most---but not all---of the accuracy indicators (\autoref{tab:validation_results}). They fall roughly in a similar range for MAPE (7.8--9.0\%), MAE (71.8--79.7 seconds), MSE (12,140--13,517 square seconds), APR (0.99--1.01), and $R^2$ (0.93), demonstrating the robustness of our model to algorithmic particulars. However, the random forest and decision tree models exhibit much lower absolute systematic prediction bias ($\delta$ = 0.45 and 0.44 seconds, respectively) than the gradient boosting and AdaBoost models do (-19.3 and -9.7 seconds, respectively). In other words, all these models have comparable mean errors---but the average random forest and decision tree predictions are statistically insignificantly different from the average true travel time, whereas gradient boosting and AdaBoost significantly under-predict it.

Therefore, we select the random forest model as our \enquote{final} travel time prediction model because it predicts better overall (than the decision tree model) and better avoids systematic under-prediction (than the gradient boosting and AdaBoost models). This final model trains an ensemble of decision trees to predict unobserved response values by essentially dividing the data into boxes then making predictions based on the means of those boxes. Its optimized hyperparameterization uses 400 decision trees, random sampling with replacement in each decision tree, a maximum decision tree depth of 10, use of all available input explanatory features at each split, a requirement of at least two samples to split at a decision node, and the default settings that require at least one sample at a leaf node and apply equal sample weighting.

Finally, we check this final model for overfitting by conducting another 5-fold cross-validation across the whole sample using the tuned hyperparameters. The resulting five MAE values (75.3, 74.5, 73.5, 73.3, 74.6) are all very close to each other, indicating that the model is not overfitted and its hyperparameters are not overly specific to the training set.

@@

\begin{table*}[tb!]
    \centering\small
    \caption{Out-of-sample prediction accuracy of our chosen model (random forest, bold) versus the initial naïve travel time model and the discarded alternative models, all validated against the corresponding true travel times. $n$ = 8,272 for each. The six accuracy indicators are (1) MAPE \%, (2) MAE in seconds, (3) MSE in square seconds, (4) difference-in-means ($\delta$) in seconds and its corresponding $t$-test's $p$-value, (5) APR, and (6) $R^2$. See methods for definitions.}\label{tab:validation_results}
    \begin{tabular}{lrrrrrrr}
        \toprule
        Model               & MAPE  & MAE    & MSE      & $\delta$ & $p$  & APR  & $R^2$ \\
        \midrule
        Initial naïve       & 21.13 & 183.55 & 48155.43 &  -182.71 & <0.01 & 0.79 & 0.74  \\
        \textbf{Random forest}       &  \textbf{8.43} &  \textbf{75.24} & \textbf{12140.14} &     \textbf{0.45} &  \textbf{0.71} & \textbf{1.01} & \textbf{0.93}  \\
        Gradient boosting   &  7.84 &  71.81 & 12572.83 &   -19.30 & <0.01 & 0.99 & 0.93  \\
        Decision trees      &  8.96 &  79.74 & 13517.03 &     0.44 &  0.73 & 1.01 & 0.93  \\
        AdaBoost            &  8.21 &  74.05 & 12416.08 &    -9.73 & <0.01 & 1.00 & 0.93  \\
        \bottomrule
    \end{tabular}
\end{table*}

Compared to the initial naïve travel time calculation, our travel time prediction model exhibits substantial out-of-sample improvement (\autoref{tab:validation_results}). Our model produces a MAPE of 8.4\%, in line with the \textasciitilde10\% MAPEs seen in the state-of-the-art travel time prediction literature, but without their extensive and expensive input data requirements. In comparison, our initial naïve travel time calculation produces a much worse MAPE of 21.1\%. Whereas the naïve travel time's MAE is 183.6 seconds, our model's predicted travel times' MAE is just 75.2 seconds---an improvement by a factor of 2.4. Similarly for MSE, our model improved by a factor of 4.0.

The $t$-test reveals significant differences between the initial naïve travel time calculations and the true travel times: the $\delta$ of -183.6 seconds corresponds to $p$ < 0.01. In other words, the naïve travel time under-predicts true travel time by over 3 minutes. However, the $t$-test reveals insignificant differences between our model's out-of-sample predictions and the true travel times: the $\delta$ of 0.45 seconds corresponds to a $p$ value of 0.71. That is, our prediction model over-predicts by less than half a second on average, which is statistically insignificantly different from zero. If we reconsider the aforementioned MAEs in light of these differences-in-means, we can see that the initial naïve travel time calculations significantly and consistently under-predict travel time, but our prediction model shows no such directional bias: its random error averages out between (smaller absolute) over-and under-prediction.

The APR offers another lens on this finding. On average across the OD pairs, the initial naïve travel time calculation under-predicts true travel time by 21\%, but our prediction model over-predicts it by just 1\%. Our model also explains more of the variation in travel time: its $R^2$ of 0.93 is substantially higher than the 0.74 $R^2$ of the initial naïve travel time calculation. In sum, each of our six accuracy indicators demonstrates that our prediction model drastically improves on naïve travel time calculations---without needing extensive, expensive input data or advanced deep learning software---and offers much more accurate travel time predictions.

@@

The planning literature shows that planners often use driving travel time prediction techniques far below the state-of-the-art due to cost, data, and technical capacity constraints. This study investigated if a better middle-ground travel time prediction method was feasible. It identified three criteria for success: (1) it should use free software and data and avoid bespoke sensor data collection; (2) it should be easy and accessible to use; (3) it should offer better accuracy than naïve predictions (e.g., minimizing Euclidean distance, network distance, street edge traversal time) and better spatial resolution than zone-to-zone travel times.

@@

This middle ground is not necessarily for computer scientists, high-end labs with top-of-the-line computing hardware, or well-resourced agencies with extensive data budgets. Rather, it empowers the scholars and practitioners who need it the most---those working on the frontlines of urban planning and policymaking. The literature shows that these scholars and practitioners often rely on simple but inaccurate routing models---such as minimizing Euclidean distance, network distance, or edge traversal time---when they lack the technical expertise, computing equipment, or extensive data needed to implement the literature's cutting-edge models. However, this study demonstrates how these naïve methods systematically under-predict real-world travel times. Our model addresses this problem by incorporating sparse open data on traffic control elements and turns into the prediction to achieve an accuracy in line with state-of-the-art methods---but without their data needs or computing requirements. In other words, our model does not replace the state-of-the-art congested-traffic methods for those with extensive resources available to implement them, but rather offers far more accurate predictions for urban planners who would otherwise fall back on simpler naïve models. Inaccurate travel times skew empirical understandings and mislead planning interventions. Our model represents an important middle-ground contribution for urban planning scholarship and evidence-informed practitioner interventions.
